rredlist
========

```{r echo=FALSE}
```{r echo=FALSE}
library("knitr")
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(if (abs(lines[1])>1) more else NULL,
            x[lines],
            if (length(x)>lines[abs(length(lines))]) more else NULL
           )
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

[![Build Status](https://travis-ci.org/ropenscilabs/rredlist.svg?branch=master)](https://travis-ci.org/ropenscilabs/rredlist)

## Install

CRAN

```{r eval=FALSE}
install.packages("rredlist")
```

Development version

```{r eval=FALSE}
devtools::install_github("ropenscilabs/rredlist")
```

```{r}
library("rredlist")
```

## High level API

High level functions do the HTTP request and parse to data to a data.frame for ease 
of downstream use. 

```{r}
rl_search('Fratercula arctica')
```

Likely a bit faster is to parse to a list only, and not take the extra data.frame parsing time

```{r output.lines=1:10}
rl_search('Fratercula arctica', parse = FALSE)
```

## Low level API

The parsing to data.frame in the high level API does take extra time. The low level API
only does the HTTP request, and gives back JSON without doing any more parsing

```{r}
rl_search_('Fratercula arctica')
```

To consume this JSON, you can use `jsonlite`

```{r}
library("jsonlite")
jsonlite::fromJSON(rl_search_('Fratercula arctica'))
```

Or other tools, e.g., `jq` via the `jqr` R client

```{r output.lines=1:20}
# devtools::install_github("ropensci/jqr")
library("jqr")
rl_search_('Fratercula arctica') %>% dot()
```

## Meta

* Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.
